Motional Quantum Ground-State Cooling Outside the Lamb-Dicke Regime -- Supplemental material

### 2017-07-17 paper.tex
More Nick Updates
Minor updates to main text, updates to SM.

\begin{document}
\title{Motional ground state Cooling Outside the Lamb-Dicke Regime}
\author{Yichao Yu}
\email{yichaoyu@g.harvard.edu}
\author{Nicholas R. Hutzler}
\altaffiliation{Present address: California Institute of Technology, Division of Physics, Mathematics, and Astronomy.  Pasadena, CA, 91125}
\author{Jessie T. Zhang}
\author{Lee R. Liu}
\author{Kang-Kuen Ni}
\email{ni@chemistry.harvard.edu}
\affiliation{Department of Chemistry and Chemical Biology, Harvard University, Cambridge, Massachusetts, 02138, USA}
\affiliation{Department of Physics, Harvard University, Cambridge, Massachusetts, 02138, USA}
\affiliation{Harvard-MIT Center for Ultracold Atoms, Cambridge, Massachusetts, 02138, USA}

@@

\begin{figure*}
  \includegraphics[height=5.2cm]{imgs/Na_RSC_schematic.pdf}
  \includegraphics[height=4.5cm]{sequence.pdf}
  \caption{(A) Energy levels and schematic of Raman sideband cooling.
    The Raman transitions have a one photon detuning $\Delta=25$ GHz from the D2 line.
    We use D1 light with $\sigma^-$ polarization to repump atoms out of $|F=2,m_F=-1\rangle$
    state to minimize heating on the atom in $|F=2,m_F=-2\rangle$ state
    (B) Geometry and polarizations of the Raman and optical pumping beams relative to the
    optical tweezer and bias magnetic field.
    (C) Schematic of the cooling sequence. The tweezer switches at 3 MHz to
    reduce light shifts during optical pumping. Each cooling cycle consists of $8$ pulses.
    The four axial pulses are addressing two neighboring cooling orders.
    The two pulses in each radial directions are either addressing two neighboring cooling orders
    or having different length on the first order when most of the population are below $n=3$
    towards then end of the cooling sequence.
    All of the Raman pulses use Blackman pulse envelopes to reduce the off resonance
    coupling to carrier and heating sidebands.
    \label{f-setup}}
\end{figure*}
\begin{figure}[b]
  \includegraphics[width=8.5cm]{imgs/fig2_raman_op.pdf}
  \caption{Optical pumping state changing probability and Raman matrix element for axial direction.
    The range plotted covers $95$\% of the initial thermal distribution.
    (A) Axial motional state distribution after one optical pumping cycle
    for different initial $n$. Due to the large Lamb-Dicke parameter,
    there is a high probability of $n$ changing.
    (B) Matrix elements for Raman transition in the axial direction showing deviation from
    $\sqrt{n}$ scaling and multiple minima for different sideband orders.
    ($k$ is the difference in wave vector of the Raman beams,
    $r$ is the coordinate in axial direction).
    \label{f-ld}}
\end{figure}
\begin{figure*}
  \includegraphics[height=4.2cm]{imgs/spectrum_r.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_r3_0.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_r3_p1.pdf}
  \caption{(A) Radial Raman sideband spectrum of first order heating, first order cooling and
    second order cooling before and after Raman sideband cooling.
    (B,C) Rabi flopping on axis 3 (B) carrier and (C) first order heating sideband
    before (red) and after (blue) cooling.
    Solid lines in (B) and (C) are theoretical calculation of the Rabi flopping.
    The blue lines corresponds to a ground state probability of $93$\% after cooling and
    the red lines corresponds to a thermal distribution of $70$ $\mu$K before cooling.
    \label{f-radial}}
\end{figure*}
\begin{figure*}
  \includegraphics[height=4.2cm]{imgs/spectrum_a1.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_a1_0.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_a1_p1.pdf}
  \caption{(A) Axial Raman sideband spectrum from first order heating to eighth order cooling
    before and after Raman sideband cooling.
    The data for the second and higher orders of cooling sidebands are taken with 150 $\mu$s
    pulse time and the rest are taken with 125 $\mu$s pulse time.
    (B,C) Rabi flopping on axial (B) carrier and (C) first order heating sideband
    before (red) and after (blue) cooling.
    Solid lines in (B) and (C) are theoretical calculation of the Rabi flopping.
    The blue and green lines corresponds to a ground state probability of $92$\% after cooling and
    the red lines corresponds to a thermal distribution of $70$ $\mu$K before cooling.
    The blue lines does not take into account the effect of decoherence due to resonance
    fluctuation. By comparing it to the green line in (C), which includes a $3$ kHz fluctuation,
    we can clearly see that this effect is the strongest for the after cooling data on
    the axial heating sideband where the Rabi frequency is the lowest.
    \label{f-axial}}
\end{figure*}

>>>>

\begin{document}
\title{Motional Ground State Cooling Outside the Lamb-Dicke Regime}
\author{Yichao Yu}
\email{yichaoyu@g.harvard.edu}
\author{Nicholas R. Hutzler}
\altaffiliation{Present address: California Institute of Technology, Division of Physics, Mathematics, and Astronomy.  Pasadena, CA, 91125}
\author{Jessie T. Zhang}
\author{Lee R. Liu}
\author{Kang-Kuen Ni}
\email{ni@chemistry.harvard.edu}
\affiliation{Department of Chemistry and Chemical Biology, Harvard University, Cambridge, Massachusetts, 02138, USA}
\affiliation{Department of Physics, Harvard University, Cambridge, Massachusetts, 02138, USA}
\affiliation{Harvard-MIT Center for Ultracold Atoms, Cambridge, Massachusetts, 02138, USA}

@@

\begin{figure*}
  \includegraphics[height=5.2cm]{imgs/Na_RSC_schematic.pdf}
  \includegraphics[height=4.5cm]{sequence.pdf}
  \caption{(A) Energy levels and schematic of Raman sideband cooling.
    The Raman transitions have a one photon detuning $\Delta=25$ GHz from the D2 line.
    We use D1 light with $\sigma^-$ polarization to repump atoms out of $|F=2,m_F=-1\rangle$
    state to minimize heating on the atom in the $|F=2,m_F=-2\rangle$ state
    (B) Geometry and polarizations of the Raman and optical pumping beams relative to the
    optical tweezer and bias magnetic field.
    (C) Schematic of the cooling sequence. The tweezer switches at 3 MHz to
    reduce light shifts during optical pumping. Each cooling cycle consists of $8$ pulses.
    The four axial pulses are addressing two neighboring cooling orders.
    The two pulses in each radial directions are either addressing two neighboring cooling orders
    or having different length on the first order when most of the population are below $n=3$
    towards then end of the cooling sequence.
    All of the Raman pulses use Blackman pulse envelopes to reduce the off resonance
    coupling to carrier and heating sidebands.
    \label{f-setup}}
\end{figure*}
\begin{figure}[b]
  \includegraphics[width=8.5cm]{imgs/fig2_raman_op.pdf}
  \caption{Optical pumping state changing probability and Raman matrix element for axial direction.
    The range plotted covers $95$\% of the initial thermal distribution.
    (A) Axial motional state distribution after one optical pumping cycle
    for different initial $n$. Due to the large Lamb-Dicke parameter,
    there is a high probability of $n$ changing.
    (B) Matrix elements for Raman transition in the axial direction showing deviation from
    $\sqrt{n}$ scaling and multiple minima for different sideband orders.
    ($k$ is the difference in wave vector of the Raman beams,
    $r$ is the coordinate in axial direction).
    \label{f-ld}}
\end{figure}
\begin{figure*}
  \includegraphics[height=4.2cm]{imgs/spectrum_r.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_r3_0.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_r3_p1.pdf}
  \caption{(A) Radial Raman sideband spectrum of first order heating, first order cooling and
    second order cooling before and after Raman sideband cooling.
    (B,C) Rabi flopping on axis 3 (B) carrier and (C) first order heating sideband
    before (red) and after (blue) cooling.
    Solid lines in (B) and (C) are theoretical calculation of the Rabi flopping.
    The blue lines corresponds to a ground state probability of $93$\% after cooling and
    the red lines corresponds to a thermal distribution of $70$ $\mu$K before cooling.
    \label{f-radial}}
\end{figure*}
\begin{figure*}
  \includegraphics[height=4.2cm]{imgs/spectrum_a1.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_a1_0.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_a1_p1.pdf}
  \caption{(A) Axial Raman sideband spectrum from first order heating to eighth order cooling
    before and after Raman sideband cooling.
    The data for the second and higher orders of cooling sidebands are taken with 150 $\mu$s
    pulse time and the rest are taken with 125 $\mu$s pulse time.
    (B,C) Rabi flopping on axial (B) carrier and (C) first order heating sideband
    before (red) and after (blue) cooling.
    Solid lines in (B) and (C) are theoretical calculation of the Rabi flopping.
    The blue and green lines corresponds to a ground state probability of $92$\% after cooling and
    the red lines corresponds to a thermal distribution of $70$ $\mu$K before cooling.
    The blue lines do not take into account the effect of decoherence due to resonance
    fluctuation. By comparing it to the green line in (C), which includes a $3$ kHz fluctuation,
    we can clearly see that this effect is the strongest for the after cooling data on
    the axial heating sideband where the Rabi frequency is the lowest.
    \label{f-axial}}
\end{figure*}

### 2017-07-17 sm.tex
Typo...
Cite julia paper
More Nick Updates
Minor updates to main text, updates to SM.

\documentclass[aps,twocolumn,secnumarabic,balancelastpage,amsmath,amssymb]{revtex4}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{color}
\usepackage{graphics}
\usepackage[pdftex]{graphicx}
\usepackage[utf8x]{inputenc}
\usepackage[colorlinks=true]{hyperref}

@@

\begin{document}
\title{Motional Quantum Ground-State Cooling Outside the Lamb-Dicke Regime -- Supplemental material}
\author{Yichao Yu}
\author{Nicholas R. Hutzler}
\author{Jessie T. Zhang}
\author{Lee R. Liu}
\author{Kang-Kuen Ni}
\email{ni@chemistry.harvard.edu}
\affiliation{Department of Chemistry and Chemical Biology, Harvard University, Cambridge, Massachusetts, 02138, USA}
\affiliation{Department of Physics, Harvard University, Cambridge, Massachusetts, 02138, USA}
\affiliation{Harvard-MIT Center for Ultracold Atoms, Cambridge, Massachusetts, 02138, USA}

@@

\subsection{Overview}

We use a computer simulation to study the importance of different effects
and guide the optimization of cooling.
The simulation uses quantum jump method\cite{Chretien2014} (or Monte-Carlo wavefunction method)
to accurately calculate the evolution within each laser pulse while different energy eigenstates
are assumed to decohere immediately on spontaneous emission of photon and on pulse boundaries.
Such approxiations are valid for simulation of the cooling and greatly simplifies the computation
at the cost of limiting the applicability of the simulation to processes that requires stronger
coherence such as Ramsey spectroscopy and rapid push-out.

A few other approximations are used in the simulation either because they are proved to have
negligible effect on the experiment or because they'll increase the complexity of the simulation
and can be optimized through other methods in the experiment.
A list of the most important approximations is,

\begin{enumerate}
\item All Raman pulses are assumed to be resonantly driving only one sideband/carrier order.\\
  We tweak the pulse shape, power and length to suppress any deviation
  from this in the experiment. We do still see deviation from this in the experiment
  but the effect is small for the short pulse we use for cooling and is the most pronounced
  for long detection pulses.
\item No heating from other sources including trap switching.\\
  We measured the heating rate and observed that it mainly comes from off-resonance scattering
  of photons from various sources which we already include so additional heating can be
  ignored.
\item Trapping frequency difference between different states are ignored.\\
  This effect is negligible for Na in the ground state due to the small fine structure.
  It is easy to include this effect if the calculation needs to be repeated on a system where
  this is not the case (e.g. directly laser cooled molecules)
\end{enumerate}

Some of the important effects that are included are,
\begin{enumerate}
\item Full 3D simulation.\\
  This includes 3D motional states, accurate 3D wavevector, and different emission pattern
  for different photon polarizations.
\item Off resonant scattering from all beams.\\
  Including Raman, optical pumping and dipole trap.
\item Optical pumping polarization imperfection.
\end{enumerate}

The optical pumping polarization and the off resonant scattering are especially important
for accurate simulation since they are the main heating source that is present
for ground motional dark state atom and therefore are the main limation on cooling performance.

\subsection{Simulation of each pulse}

Each pulse in general has a coherent part from the Raman drive (zero for optical pumping pulse)
and an incoherent part from scattering (optical pumping or off-resonance scattering).
This nicely fits the general formula of the quantum jump method. Due to the large number of
states that we need to take into account ($30-150$ states in each of the three axis and
$8$ hyperfine states, giving a total of $10^6$ states),
the generic wavefunction propagation is very inefficient.
Therefore, this is further simplified by reducing the number of states to be kept track of during
the coherent propagation and apply quantum jump using a set of reduced jump operator that
describes the total jump probability from a certain states (sum over all final states).
The coherent propagation is then accurately solved in order to avoid having to use a time step
much shorter than the scattering rate (up to hundreds of kHz)\footnote{\href{https://goo.gl/EZ13wt}{note about modification to the quantum jump method (https://goo.gl/EZ13wt)}}.

In order to simulate a pulse, we first start with the coherent propagation that includes the
modification from the jump operators as in quantum jump method.
When a jump happens, we pick a final internal state to jump to based on the rates ratio
of different scattering sources and their branching ratios.
This process also determines the polarization and emission pattern of the photon and
a particular wave vector for the emitted photon is then selected as the next step.
The recoil is not fixed so the final motional state of the atom can now be selected.
This process is then repeated on the same pulse until the pulse time is used up before
we start to simulate the next pulse.

\subsection{Measurements}

We apply ``virtual'' measurements on each results of the cooling which we then average to
obtain the expected value of qualities that we are interested in.
Due to the decoherence approximation, the measurement has to commute with the Hamiltonian.
Particularly useful measurements used in our simulation are hyperfine state distribution,
Average motional energy, loss rate (where an atom is assumed to be lost if its motional energy
goes above a certain threshold at any time) probability of being in certain states
(e.g. ground state).

\subsection{Sample results}

\begin{figure}
  \includegraphics[width=8.5cm]{imgs/simcool_no_scatter.pdf}
  \caption{Simulated ground state population as a function of cooling cycles without
    technical imperfections. The ground state probability reaches $>99\%$. \label{f-no-scatter}}
\end{figure}

\begin{figure}
  \includegraphics[width=8.5cm]{imgs/simcool_real.pdf}
  \caption{Simulated ground state population as a function of cooling cycles including effect of
    off-resonant scattering from all beams and OP misalignment.
    For this particular sequence, the ground state probability saturates $\approx88\%$.
    \label{f-real}}
\end{figure}

When not considering the off-resonant scattering from the Raman beams and the tweezer as well
as optical pumping polarization misalignment, the simulation shows that there isn't a cooling
limit despite the large Lamb-Dicke parameter and high initial temperature (Fig \ref{f-no-scatter}).
However, when these effects are taken into account, the ground state probability can saturate at
a lower value (Fig \ref{f-real}) and require much more careful tweaking.

\subsection{Code}

The simulation is written in Julia. All code are available under GPLv3 license at
\footnote{\href{https://goo.gl/WLJwJp}{Library code (https://goo.gl/WLJwJp)}} and
\footnote{\href{https://goo.gl/XXVreq}{Top level code (https://goo.gl/XXVreq)}}

>>>>

\documentclass[aps,twocolumn,secnumarabic,amsmath,amssymb]{revtex4}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{color}
\usepackage{graphics}
\usepackage[pdftex]{graphicx}
\usepackage[utf8x]{inputenc}
\usepackage[colorlinks=true]{hyperref}

@@

\begin{document}
\title{Motional Quantum Ground-State Cooling Outside the Lamb-Dicke Regime -- Supplemental material}
\author{Yichao Yu}
\author{Nicholas R. Hutzler}
\altaffiliation{Present address: California Institute of Technology, Division of Physics, Mathematics, and Astronomy.  Pasadena, CA, 91125}
\author{Jessie T. Zhang}
\author{Lee R. Liu}
\author{Kang-Kuen Ni}
\email{ni@chemistry.harvard.edu}
\affiliation{Department of Chemistry and Chemical Biology, Harvard University, Cambridge, Massachusetts, 02138, USA}
\affiliation{Department of Physics, Harvard University, Cambridge, Massachusetts, 02138, USA}
\affiliation{Harvard-MIT Center for Ultracold Atoms, Cambridge, Massachusetts, 02138, USA}

@@

%\subsection{Overview}

We use a computer simulation to study the importance of different effects
and guide the optimization of Raman sideband cooling.
The simulation uses the quantum jump (or Monte-Carlo wavefunction) method~\cite{Chretien2014} 
to accurately calculate the evolution within each laser pulse where different energy eigenstates
are assumed to decohere immediately upon spontaneous emission and on pulse boundaries.
Such approximations are valid and greatly simplifies the computation
at the cost of limiting the applicability of the simulation to processes that requires stronger
coherence, such as Ramsey spectroscopy or rapid push-out of the atoms from the trap.

A number of other simplifying approximations, listed below, are also used in the simulation.  These either have a demonstrably negligible effect on the experiment, or result in a significant reduction of complexity for an effect that can be optimized more easily in the experiment.

\begin{enumerate}
\item \emph{All Raman pulses resonantly drive only one sideband/carrier order.}  We tweak the pulse shape, power and length as discussed in the text, for example by using Blackman pulses, to reduce coupling to other orders. We do see deviation from this approximation in the experiment, though mostly in the long detection pulses, and it is less of a concern for the short cooling pulses.
\item \emph{No heating from other sources.}\\
  We measured the heating rate and found that it results mainly from off-resonant scattering
  of photons, which is included in the simulation.  In particular, we can ignore heating from the switching trap.
\item \emph{Same trapping frequency for all states.}\\
  This effect is negligible for Na in the ground state due to the small fine structure splitting in the excited state.
  It is easy to include this effect if the calculation needs to be repeated on a system where
  this is not the case (e.g. directly laser cooled molecules)
\end{enumerate}

Some effects that are important to include are all three dimensions (3D motional states, accurate 3D wavevector, and different emission pattern for different photon polarizations), off-resonant scattering from all laser beams, and optical pumping polarization imperfection.  The optical pumping polarization and off-resonant scattering are especially important
for accurate simulation since they are the main limitation on cooling performance.

%\subsection{Simulation of each pulse}

In general, each pulse has a coherent part from the Raman drive (which is absent for the optical pumping pulses)
and an incoherent part from scattering (optical pumping or off-resonant).
This nicely fits the general approach of the quantum jump method. Due to the large number of
states that we need to take into account ($30-150$ motional per axis and
$8$ hyperfine, giving a total of $10^6$),
 generic wavefunction propagation is very inefficient.
Therefore, we further simplify by reducing the number of states tracked during
the coherent propagation and apply quantum jumps using a set of reduced jump operator that
describes the total jump probability from a certain states (summed over all final states).
The coherent propagation is then accurately solved in order to avoid having to use a time step
much shorter than the scattering rate (up to hundreds of kHz)\footnote{\href{https://goo.gl/EZ13wt}{note about modification to the quantum jump method (https://goo.gl/EZ13wt)}}.

In order to simulate a pulse, we first start with the coherent propagation that includes the
modification from the jump operators as in quantum jump method.
When a jump happens, we pick a final internal state to jump to based on the relative rates
of different scattering sources and their branching ratios.
This process also determines the polarization and emission pattern of the photon and
a particular wave vector for the emitted photon is then selected as the next step.
The recoil is now fixed so the final motional state of the atom can now be selected.
This process is then repeated on the same pulse until the pulse time is used up before
we start to simulate the next pulse.

%\subsection{Measurements}

We apply ``virtual'' measurements on the results of each simulated cooling sequence which we then average to
obtain the expected value of experimental quantities.
Due to the decoherence approximation, the measurement has to commute with the Hamiltonian.
Particularly useful measurements used in our simulation are hyperfine state distribution,
average motional energy, loss rate (where an atom is assumed to be lost if its motional energy
goes above a certain threshold at any time) probability of being in certain states
(e.g. ground state).

\begin{figure}[t]
  \includegraphics[width=8.5cm]{imgs/simcool_no_scatter.pdf}
  \caption{Simulated ground state population as a function of cooling cycles without
    technical imperfections. The ground state probability reaches $>99\%$. \label{f-no-scatter}}
\end{figure}
\begin{figure}[t]
  \includegraphics[width=8.5cm]{imgs/simcool_real.pdf}
  \caption{Simulated ground state population as a function of cooling cycles including effect of
    off-resonant scattering from all beams and OP misalignment.
    For this particular sequence, the ground state probability saturates to $\approx88\%$.
    \label{f-real}}
\end{figure}

If off-resonant scattering and optical pumping polarization misalignment are ignored, the simulation shows that there isn't a cooling
limit despite the large Lamb-Dicke parameter and high initial temperature (Fig \ref{f-no-scatter}).
However, when these effects are taken into account, the ground state probability can saturate at
a lower value (Fig \ref{f-real}) and require much more careful tweaking.

%\subsection{Code}

The simulation is written in Julia\cite{Bezanson2017}.
All code is available under the GPLv3 license at
\footnote{\href{https://goo.gl/WLJwJp}{Library code (https://goo.gl/WLJwJp)}} and
\footnote{\href{https://goo.gl/XXVreq}{Top level code (https://goo.gl/XXVreq)}}
